---
title: "Bayesian lab"
author: Ben Bolker
bibliography: "../glmm.bib"
date: "`r format(Sys.time(), '%d %B %Y ')`"
---

```{r pkgs}
library(MCMCglmm)
library(brms)
library(coda)
library(lattice)
library(reshape2) ## for melt()
## devtools::install_github("mjskay/tidybayes")
## library(tidybayes) ## optional
library(ggplot2); theme_set(theme_bw())
```

```{r data}
load("../data/starling.RData")   ## loads "dataf"
```

### Analyze with `MCMCglmm`

(optional)

`MCMCglmm` does a Bayesian analysis, which can be useful for more
flexible inference, and has some more flexible covariance structures etc..
For basic models, the syntax is very similar to `lme4`/`nlme`.

```{r mcmcglmm,message=FALSE,cache=TRUE}
mcmcglmm1 <- MCMCglmm(stmass~mnth*roostsitu,
                      random=~subject,
                      data=dataf,
                      verbose=FALSE)
```
We use `verbose=FALSE` to turn off the progress messages, which would be ugly in this document but are generally useful.

* Compare the results (use `summary()`: printing out the a raw `MCMCglmm` model is ugly).

For MCMC approaches, it is your responsibility to check that the chain(s) are well-behaved.

Try this:
```{r mcmcplot1,fig.keep="none"}
library("coda")
xyplot(as.mcmc(mcmcglmm1$Sol),layout=c(2,4))
```

These look the way they should (featureless
"white noise" - no trends or slow variation).

You can plot the distributions:
```{r plotdens,fig.keep="none"}
densityplot(mcmcglmm1$Sol)
```
but I prefer *violin plots* for this case:
```{r plotviolin,fig.keep="none"}
md <- reshape2::melt(as.matrix(mcmcglmm1$Sol)) ## change matrix to long form
ggplot(subset(md,Var2!="(Intercept)"),
       aes(Var2,value))+geom_violin(fill="grey")+
    geom_hline(yintercept=0,lty=2)+
    coord_flip()
```    
