---
title: "Mixed model lab #2 (GLMMs)"
author: Ben Bolker
date: "`r format(Sys.time(), '%H:%M %d %B %Y ')`"
---

```{r pkgs,message=FALSE}
library(lme4)
library(glmmTMB)
library(aods3)
library(glmmTMB)
library(broom)
library(broom.mixed)
library(dotwhisker)
```

## CBPP data

Starting again with the "basic" CBPP model:

```{r cbpp}
data("cbpp",package="lme4")
g1 <- glmer(incidence/size~period+(1|herd),
      data=cbpp,
      weights=size,
      family=binomial)
```

What happens if we try to fit the *maximal* model?

```{r maxvars,error=TRUE}
g1max <- update(g1,.~.-(1|herd) + (period|herd))
```

Use `glmerControl()` to override the warning ...

```{r maxvars2}
g1max <- glmer(incidence/size~period+(period|herd),
      data=cbpp,
      weights=size,
      family=binomial,
      control=glmerControl(check.nobs.vs.nRE="warning"))
VarCorr(g1max)
eigen(VarCorr(g1max)[[1]])$values
```

Look at `?convergence` ...

```{r allfit,cache=TRUE}
source("../R/allFit.R")
af <- allFit(g1max)
```

We can use `summary(af)` to compare all of these fits.

## gopher tortoise data

How do we reduce this to something that will work?

```{r}
load("../data/gopherdat2.RData")
head(Gdat)
```

```{r eval=FALSE}
Gdat$obs <- factor(seq(nrow(Gdat)))
glmer(shells~prev+offset(log(Area))+(1|year)+(1|Site),
      family=poisson,data=Gdat)
glmer(shells~prev+offset(log(Area))+year+(1|Site),
      family=poisson,data=Gdat)

glmer(shells~prev+offset(log(Area))+year+Site+
          (1|obs),
      family=poisson,data=Gdat)
glmmTMB(shells~prev+offset(log(Area))+year+Site+
          (1|obs),
      family=poisson,data=Gdat)
```

## Arabidopsis data

Try full model (except use only a simple fixed effect for the top-level region, `reg`, which has only 3 levels):

```{r banta1,cache=TRUE}
load("../data/Banta.RData")
t0 <- system.time(
    mp1 <- glmer(total.fruits ~ nutrient*amd +
                 reg + rack + status +
                 (amd*nutrient|popu)+
                 (amd*nutrient|gen),
             data=dat.tf,
             family="poisson")
)
```

Inspect for singularity/overdispersion:

```{r banta_chk}
eigen(VarCorr(mp1)$gen)$values
eigen(VarCorr(mp1)$popu)$values
deviance(mp1)/df.residual(mp1) ## !!
aods3::gof(mp1)
```

Add observation-level random effect (switch optimizer to BOBYQA)

```{r banta2,cache=TRUE}
dat.tf$obs <- 1:nrow(dat.tf)
t1 <- system.time(
    mp1X <- update(mp1,
                   . ~. + (1|obs),
                   control=glmerControl(optimizer="bobyqa"))
)
```

This model has 28 parameters and takes `r round(t1["elapsed"])` seconds ... Is `glmmTMB` faster?

```{r glmmTMBfit,cache=TRUE}
t2 <- system.time(
    mp1g <- glmmTMB(total.fruits ~ nutrient*amd +
                        reg + rack + status +
                        (amd*nutrient|popu)+
                        (amd*nutrient|gen)+
                        (1|obs),
                    data=dat.tf,
                    family="poisson")
)
```

Better (`r round(t2["elapsed"])`), although 
`glmmTMB` doesn't handle singular fits quite as gracefully ...

```{r reduce1,cache=TRUE}
mp1X2 <- update(mp1X,
                . ~ . - (amd*nutrient|popu)
                - (amd*nutrient|gen)
                + (amd+nutrient|popu)
                + (amd+nutrient|gen))
```

Ugh, looks even worse ...

```{r showreduce}
VarCorr(mp1X2)
```

Use `lmer_alt` to split factor variable into separate terms ...
```{r X3,cache=TRUE}
mp1X3 <- afex::lmer_alt(total.fruits ~ nutrient*amd +
                            reg + rack + status +
                            (amd*nutrient||popu)+
                            (amd*nutrient||gen)+
                            (1|obs),
                        data=dat.tf,
                        family="poisson")
```

Strip down further:
```{r X4,cache=TRUE}
mp1X4 <- update(mp1X2,
                . ~ . - (amd+nutrient|popu)
                - (amd+nutrient|gen)
                + (nutrient|popu)
                + (1|gen))
```

Still correlation=+1, reduce further:

```{r X5,cache=TRUE}
mp1X5 <- afex::lmer_alt(total.fruits ~ nutrient*amd +
                            reg + rack + status +
                            (1|gen)+
                            (1+nutrient||popu)+
                            (1|obs),
                        data=dat.tf,
                        family="poisson")
```
