---
title: "Generalized linear mixed models in R: nitty-gritty"
author: Ben Bolker
bibliography: "../glmm.bib"
date: "`r format(Sys.time(), '%d %B %Y ')`"
---

```{r pkgs,message=FALSE}
library(lme4)
library(broom)
library(broom.mixed)
library(dotwhisker)
```
        
## Overview

Most aspects of GLMMs are carried over from LMMs (random effects) and GLMs (families and links).

## integration methods

You do have to decide on an approximation method.

- penalized quasi-likelihood
    - fastest/least accurate
	- uses quasi-likelihood (no AIC etc.?)
- Laplace approximation
    - fast, flexible compromise
- adaptive Gauss-Hermite quadrature
    - slowest
    - most accurate
    - most limited

In `lme4`, use the `nAGQ=` argument; `nAGQ=1` (default) corresponds to Laplace approximation

## integration rules

- it only really matters in relatively extreme cases (small numbers of binary obs per group)
- use the most accurate method that's available and feasible

## Laplace-approximation diagnostics

```{r lapl,fig.width=6,fig.height=5}
library(lattice)
aspect <- 0.6
xlab <- "z"; ylab <- "density"; type <- c("g","l"); scaled <- FALSE
mm <- readRDS("../data/toenail_lapldiag.rds")
print(xyplot(y ~ zvals|id, data=mm,
             type=type, aspect=aspect,
             xlab=xlab,ylab=ylab,
             as.table=TRUE,
             panel=function(x,y,...){
    if (!scaled) {
        panel.lines(x, dnorm(x), lty=2)
    } else {
        panel.abline(h=1, lty=2)
    }
    panel.xyplot(x,y,...)
}))
```

## comparing integration methods

```{r compare}
g1 <- glmer(incidence/size ~ period + (1|herd),
            family=binomial,
            data=cbpp,
            weights=size)
g2 <- update(g1,nAGQ=5)
g3 <- update(g1,nAGQ=10)
g4 <- MASS:::glmmPQL(incidence/size ~ period,
                     random = ~1|herd,
                     data=cbpp,
                     family=binomial,
                     weights=size)
dwplot(list(Laplace=g1,AGQ5=g2,AGQ10=g3,glmmPQL=g4))
```      
      
## overdispersion

- reminder: "too much" variance
- only applies to families with *estimated* variance:  
e.g. Poisson, binomial ($N>1$)
- sum^2 of Pearson residuals / residual degrees of freedom
- e.g. `aods3::gof()`

## dealing with overdispersion

- quasi-likelihood
- observation-level random effects [@elston_analysis_2001]
- overdispersed distributions  
(negative binomial, beta-binomial, etc.)  
`glmmTMB`/`brms` [@brooks_modeling_2017]

## other diagnostics

- assumption of Normal residuals may not be very good
- simulated residuals (`DHARMa`)

## zero-inflation

- `glmmTMB` etc.
- Owls example

## Complete separation

- penalized regression
- add priors!

